---
title: "Antimicrobial Use Point Prevalence Survey"
subtitle: "Tamale Teaching Hospital – Hospital-Wide Assessment of Prescribing Practices"
author: "TTH Antimicrobial Stewardship Committee"
date: "November 7, 2025"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-fold: true
    code-overflow: wrap
execute:
  warning: false
  message: false
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(scales)
library(patchwork)
library(plotly)
library(DT)
library(RColorBrewer)
library(plotly)
library(here)

# Dataframes
over_prev <- read_csv("../over_prev.csv")
daily_prev <- read_csv("../daily_prev.csv")
main_department_p_dat <- read_csv("../main_department_p_dat.csv")
sub_department <- read_csv("../sub_department_p_dat.csv")
prop_tth_wards <- read_csv("../prop_tth_wards.csv")
type_department <- read_csv("../type_department_p_dat.csv")
daily_average <- 65.1
sub_bmc_dat <- read_csv("../sub_bmc_dat.csv")
wards_dat <- read_csv("../prop_tth_wards.csv")
indication <- read_csv("../indication.csv")
top_antimicrobials <- read_csv("../top_10_antimicrobial_name.csv")
aware_watch <-read_csv("../aware.csv")
guideline <- read_csv("../guideline_compliance.csv")
reason_inc <- read_csv("../documneted_indicated.csv")
stop_review_documneted <- read_csv("../stop_review_documneted.csv")
route_admin <- read_csv("../route_admin.csv")
treat_data <-  read_csv("../treatment.csv")
treat_biodata <- read_csv("../biomarker_based_trt.csv")
culture_st_data <- read_csv("../culture_lab_status.csv")
culture_st_compl <- read_csv("../culture_lab_complince.csv")
treatment <- read_csv("../treatment.csv")
```


:::{.panel-tabset}

## Online Version
Full report content in HTML format...

## PDF Version
[Download PDF Report](/reports/ams-report.pdf){ .btn .btn-primary .btn-lg }

<iframe src="/index.pdf" width="100%" height="600px" style="border: none; margin-top: 1rem;">
</iframe>

:::


## Executive Summary

The Point Prevalence Survey (PPS) conducted in May 2024 assessed antimicrobial prescribing patterns across Tamale Teaching Hospital. Key findings highlight critical gaps in stewardship practices:

```{r}
#| echo: false
#| label: tbl-summary
#| tbl-cap: "Key PPS Findings and Interpretation"

summary_data <- data.frame(
  Metric = c("Patients Surveyed", "Antimicrobial Use Prevalence", 
             "Guideline Adherence", "Culture Before Treatment"),
  Value = c("427", "65.1%", "45.2%", "14.7%"),
  Interpretation = c("Full hospital coverage", "High usage; requires review", 
                     "Below target; training needed", "Critical gap; diagnostic stewardship needed")
)

kable(summary_data, align = c('l', 'c', 'l'))
```

## Survey Overview & Methodology

### Objective

To evaluate antimicrobial prescribing practices, identify stewardship gaps, and inform quality improvement strategies.

### Design & Setting

-   **Type:** Point Prevalence Survey
-   **Date:** May 2024\
-   **Location:** Tamale Teaching Hospital
-   **Departments Surveyed:** Internal Medicine, Surgery, Pediatrics, Obstetrics & Gynecology, Polyclinic, Trauma & Orthopaedics, DEENT, Accident & Emergency
-   **Tool Used:** Standardized PPS Form

```{r}
#| label: prevalence
#| echo: false
#| message: false
#| warning: false

patients_surveyed <- over_prev |> pull(num_admitted_patients)
patients_on_antimicrobials <- over_prev |> pull(num_antimicroial_patients)

# Overall prevalence
overall_prevalence <- over_prev |> pull(prop_antimicrbial_patient)
```

## Key Findings

### Antimicrobial Use Prevalence

**Overall Rate:** **`r overall_prevalence`%**, with `r sum(patients_on_antimicrobials)` out of `r sum(patients_surveyed)` surveyed patients receiving at least one antimicrobial agent. Visual charts show variation in use across departments, with some departments exceeding **`r overall_prevalence`%** the hospital wide rate.

```{r}
#| label: submc-breakdown
#| fig-cap: "Antimicrobial Use Prevalence by Department"
#| echo: false

# Add percentage column
sub_data <- sub_bmc_dat %>%
  mutate(percentage = round((num_antimicroial_patients / num_admitted_patients) * 100, 1),
         label = paste0(num_antimicroial_patients, " (", percentage, "%)"))

ggplot(sub_data, aes(x = fct_reorder(sub_bmc, percentage), y = percentage)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = paste0(percentage, "%")), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(x = "Department", y = "Prevalence (%)") +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 90))
```

### Indications for Prescribing

**Infection Types:** Distribution includes respiratory, urinary, surgical site, and bloodstream infections.

```{r}
#| label: indications
#| echo: false
#| message: false
#| warning: false

# Add percentage column
indication_data <- indication %>%
  mutate(percentage = round((count / 598) * 100, 1),
         label = paste0(count, " (", percentage, "%)"))

```

### Indications for Prescribing

**Infection Types:** Distribution of indicators includes Community Acquired Infections, Surgical Prophylaxis and Hospital Acquired Infections.

```{r}
#| label: infection-types
#| fig-cap: "Distribution of Infection Types"
#| echo: false

ggplot(indication_data, aes(x = reorder(indic, percentage), y = percentage, fill = indic)) +
  geom_col(fill = "coral",alpha = 0.8) +
  geom_text(aes(label = paste0(percentage, "%")), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(x = "Infection Type", y = "Percentage (%)") +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 45))
```

**Empirical vs. Targeted Therapy:** Majority of prescriptions were empirical, reflecting low culture submission rates.

```{r}
#| echo: false
#| label: fig-therapy-type
#| fig-cap: "Distribution of Empirical vs Targeted Therapy"

ggplot(treatment, aes(x = "", y = percentage, fill = trt_st)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(fill = "Therapy Type") +
  theme_void() +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_stack(vjust = 0.5)) +
  scale_fill_manual(values = c("Empirical Therapy" = "lightblue", 
                               "Targeted Therapy" = "lightcoral"))
```

### Antimicrobial Classes & Agents

**Top Agents:** Ceftriaxone, Metronidazole, and cefuroxime were most frequently prescribed. With a balanced 1:1 AWaRe Ratio

```{r}
#| label: top-agents
#| fig-cap: "Most Frequently Prescribed Antimicrobial Agents"
#| echo: false

top_antimicrobials <- top_antimicrobials %>%
  mutate(percentage = round((n / 598) * 100, 1),
         label = paste0(n, " (", percentage, "%)"))


ggplot(top_antimicrobials, aes(x = fct_reorder(antimicrobial_name, n), y = n, fill = aware)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(x = "Antimicrobial Agent", y = "Frequency", fill = "WHO AWaRe Category") +
  theme_minimal() +
  scale_fill_manual(values = c("ACCESS" = "#2E8B57", "WATCH" = "#FFA500", "NA" = "#DC143C")) +
  scale_y_continuous(limits = c(0, 140))

```

**WHO AWaRe Classification:** Majority of agents fall under the 'Watch' category, indicating potential for resistance.

-   **The Access-to-Watch ratio** is nearly 1:1, which is not ideal. WHO recommends a higher proportion of Access antibiotics to minimize resistance risks.

-   This balance suggests a need for stewardship interventions to promote Access agents and reduce reliance on Watch antibiotics unless clinically justified.

```{r}
#| label: aware-classification
#| fig-cap: "Distribution of Antimicrobials by WHO AWaRe Classification"
#| echo: false

# Add percentage column
aware_watch_data <- aware_watch %>%
  mutate(percentage = round((count / 598) * 100, 1),
         label = paste0(count, " (", percentage, "%)"))

# Bar chart for WHO AWaRe Classification Distribution
ggplot(aware_watch_data, aes(x = reorder(aware_st, -count), y = percentage, fill = aware_st)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = percentage), vjust = -0.5, size = 4) +
  labs(x = "Classification", y = "Percentage (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.position = "none") +
  scale_fill_manual(values = c("Access" = "#2E8B57", 
                                "Watch" = "#FFA500", 
                                "N/A" = "#DC143C"))
```

### Quality Indicators Compliance

-   **Stop/Review Date** (56.2%): Moderate compliance; over half of prescriptions had a review date, but improvement is needed to ensure timely reassessment.
-   **Guideline Adherence** (45.2%): Low adherence; indicates a need for training and regular audits to promote evidence-based prescribing.
-   **Culture Before Treatment** (14.7%): Critically low; reflects poor diagnostic stewardship and over-reliance on empirical therapy.
-   **Biomarker-Based Treatment** (34.1%): Limited use; suggests potential to expand diagnostic support for antimicrobial decisions.
-   **Reason for Indication Documented** (40.8%): Weak documentation; undermines accountability and hinders audit effectiveness.

```{r}
#| label: quality-indicators
#| echo: false
#| message: false
#| warning: false

quality_data <- data.frame(
  Indicator = c("Stop/Review Date", "Guideline Adherence", "Culture Before Treatment",
     "Biomarker-based Treatment", "Reason for Indication"
    ),
  Compliance = c(56.2,  45.2, 14.7, 34.1, 40.8)
)

```

```{r}
#| label: quality-chart
#| fig-cap: "Compliance with Antimicrobial Stewardship Quality Indicators"
#| echo: false

ggplot(quality_data, aes(x = reorder(Indicator, Compliance), y = Compliance)) +
  geom_col(aes(fill = Compliance), alpha = 0.8, width = 0.7) +
  geom_text(aes(label = paste0(Compliance, "%"), y = Compliance/2), 
            color = "white", size = 3, fontface = "bold") +
  labs(x = "Quality Indicator", y = "Compliance Rate (%)") +
  coord_flip() +
  theme_minimal() +
  scale_fill_gradient(low = "#FF6B6B", high = "#2E8B57") +
  theme(legend.position = "none")
```

### Microbiology & Resistance Data

**Culture Submission Rates by Ward:** Lowest in high-prescribing wards such as Surgery and A&E.\
**Resistance Trends:** Data not available; future surveys should include antibiogram analysis.

### Departmental Contributions to Antimicrobial Consumption

**Highest Rates**:

-Internal Medicine (23.1%) and Surgery (22.5%) lead in culture submission, but still far below ideal (\>80%).

**Moderate Rates**:

-TTH overall (14.7%), Paediatrics (14.1%), and Obstetrics & Gynaecology (12.2%) show low compliance.

**Critical Gaps**:

-   Accident & Emergency (5.9%) and Orthopaedics & Trauma (5%) have very poor submission rates.
-   DEENT and Polyclinic (0%) indicate no cultures submitted at all, which is alarming.

```{r}
#| label: culture-rates
#| fig-cap: "Microbiological Culture Submission Rates Before Treatment by Ward"
#| echo: false
ggplot(culture_st_compl, aes(x = reorder(sub_bmc, -percentage), y = percentage, fill = sub_bmc)) +
  geom_col(alpha = 0.9, width = 0.7) +
  geom_text(aes(label = paste0(percentage, "%")), 
            vjust = -0.3, size = 4, color = "black", fontface = "bold") +
  labs(
    title = "Culture Submission Before Antimicrobial Treatment",
    x = "Ward",
    y = "Submission Rate (%)"
  ) +
  scale_fill_manual(
    values = setNames(
      colorRampPalette(brewer.pal(8, "Dark2"))(length(unique(culture_st_compl$sub_bmc))),
      unique(culture_st_compl$sub_bmc)
    ),
    name = "Ward"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    legend.position = "none"
  ) +
  ylim(0, max(culture_st_compl$percentage, na.rm = TRUE) + 10) 


```

### Top Wards/Unit contributing to Most of AM Consumption

-   **Left Side (Red Bars)**: These wards account for the largest share of antimicrobial use. The first three represent the top contributors i.e ICU, MICU, and NICU.

-   **Cumulative curve** reaches \~80% by around 22 out of 36 wards, contributes most of the antimicrobial consumption.

-   **Right Side (Gray Bars)**: Wards with lower antimicrobial use have minimal impact on overall consumption.

```{r}
#| label: benchmark-chart
#| fig-cap: "Top Wards/Unit contributing to Most of AM Consumption"
#| echo: false
#| warning: false


# Original ggplot
ggplot(wards_dat, aes(x = reorder(department_name, -prop_antimicrbial_patient))) +
  geom_bar(aes(y = prop_antimicrbial_patient, fill = top80,
               text = paste0("Ward: ", department_name,
                             "<br>Proportion: ", percent(prop_antimicrbial_patient, accuracy = 0.1),
                             "<br>Cumulative: ", percent(cumulative, accuracy = 0.1))),
           stat = "identity", color = "black") +
  geom_line(aes(y = cumulative), group = 1, color = "#FF6600", size = 1.2) +
  geom_point(aes(y = cumulative), color = "#FF6600", size = 2)  +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "#267719") +
  geom_hline(yintercept = 1.0, linetype = "dotted", color = "gray40") +
  annotate("text", x = length(wards_dat$department_name), y = 1.02, label = "100% cumulative", color = "gray40", size = 3) +
  scale_fill_manual(values = c("TRUE" = "#B22222", "FALSE" = "gray80")) +
  scale_y_continuous(labels = percent_format(), sec.axis = sec_axis(~ ., name = "Cumulative %")) +
  labs(
    x = "Ward",
    y = "Proportion of Antimicrobial Use"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 55, hjust = 1),
    legend.position = "none"
  )
```

## Recommendations

### Short-Term Actions

-   Conduct targeted training on guideline-based prescribing
-   Promote culture-before-treatment campaigns

### Medium-Term Goals

-   Integrate stewardship prompts into EHR workflows
-   Establish AMS champions in each department

### Long-Term Strategy

-   Develop a hospital-wide antibiogram
-   Institutionalize quarterly PPS audits

------------------------------------------------------------------------

<br>
<br>

:::{.team-footer}
**Prepared by:** TTH AMS Committee

**Team:**

- Dr. Ana Maria Simono Charadan [Consultant & Chair-person]
- Dr. Samuel Yao Alomatu [Infectious Diseases Specialist]
- Dr. Elvis Duorina, [Specialist Pharmacist]
- Mr. Alhassan M. Shamudeen, [Head of Research and Development]
- Dr. David Eklu Zeyeh, MLS, [Medical Microbiologist]
- Dr. Anthony Kwaw [Specialist Pharmacist]
- Dr. Prince Parku [Specialist Pharmacist]
- Mr Mohammed Lukman [Infectious Diseases Nurse]
- Mr. Ameko Asiwome [Health Information]
:::


